|

### 2023-05-31 manuscript/paper.Rmd
Updated a bunch of bib engtries and added Joel's text to the document.

---
title: |
  |  \Large Stickle!
author: | 
  | \large Matthew Stuart \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`mstuart1@luc.edu`](mailto:mstuart1@luc.edu) \vspace{-1.1mm}
  |
  | \large Akhil Ghosh \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1.1mm}
  |
  | \large Yoel E. Stuart \vspace{-1.1mm}
  | \normalsize Department of Biology \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1.1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1.1mm}
abstract: |
  | Evewryone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

# Introduction {#sec:intro}
Sexual Dimorphism is interesting because......

@@

Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.  However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.  For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.  We view the sex of the fossils as missing data and use multiple imputation (@little2002statistical) to impute the sex of the fossils using the modern stickleback fish with observed sex to model the relationship between sex and observed phenotypes.  Once sex is imputed, an Ornstein–Uhlenbeck (OU) model is fit using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes (e.g. length, vertebrae count?, etc).

@@

# Data {#sec:data}

```{r}
stickle <- read_csv('stickle_revamped.csv')

@@

The data used here consists of a total of `r sum(!is.na(stick$gender))` extant specimens with known sex all collected in the last 30 years.  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

In addition there are `r sum(is.na(stick$gender))` fossil specimens from approximately 10.3 million years ago with unknown sex over 18 time periods spaced about 1000 years apart. Table 1 shows the sample size at each of the 18 time periods.  There are at least 22 specimens at each time period with a high of 67 specimens in period 7.

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
\begin{table}[ht]
\centering
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
\label{tab1}
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\end{table}

@@

We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.  The imputation algorithm is run to obtain a total of $M = 100$ completed datasets.

## Completed Data Model 
For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).

Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume
\begin{align}
{X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\
{X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align}
where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.  We further set \begin{align}
\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.  We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUprocess) where we assume $u_{k,gT}$ have a marginal mean of 0. Because each time period is ~1000 years, we will discretize the OU process without loss of generality.  For $t = 2,\ldots,T$, we assume
\begin{align}
u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
\label{eq:u_ar1}
\end{align}

@@

For the discrete phenotypes, we will assume
\begin{align}
{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
\label{eq:Y}
\end{align}
for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set
\begin{align}
\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
\label{eq:lambda}
\end{align}
We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume
\begin{align}
v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
\label{eq:v_ar1}
\end{align}
and
\begin{align}
v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
\label{eq:v1}
\end{align}

@@

# Future work and conclusions {#sec:conclusions}

@@

Stickle!

>>>>

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Allison Ozark$^1$, Akhil Ghosh$^1$, Raheyma Siddiqui$^1$, Matthew Stuart$^1$,
  | \large  Samantha Swank$^1$, Michael A. Bell$^2$, Gregory J. Matthews$^1$, and Yoel E. Stuart$^1$ \vspace{-1.1mm}
  |
  | \large $^1$ Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  |
  | \large $^2$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  |
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

# Introduction {#sec:intro}

Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed (reviewed in @Herrmann2021). Niche changes in habitat and resource use should then drive phenotypic evolution as the population adapts to new conditions (@Herrmann2021). For example, resource use expansion should result in trophic trait evolution (cite; e.g.,). Habitat expansion should result in locomotor change (cite; e.g.,). And release from enemies should result in reduction of costly defense traits (cite; e.g.,).

Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use and associated phenotypes (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Such character displacement between the sexes is therefore one explanation for sexual dimorphism (de Lisle papers).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution lineage of the fossil stickleback fish (Gasterosteus doryssus). The sex of the individual fossils, which is unobserved in practice, was imputed using multiple imputation by chained equations (MICE) (@MICE) with observed data used to train the imputation model based on extant stickleback with known sex.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense. The fossil sequence captures a bout of adaptive evolution following ecological release from predators, as follows. The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; Roesti et al. 2023). The lineage thus adaptively evolved armor loss and reduction in several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al. in prep), and tooth wear data suggest that the population also began eating more planktonic prey, expanding to an open water lifestyle (Purnell et al. 2007).

Sexual Dimorphism is interesting because......

@@

@@

# Methods and Materials 
<!-- Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.  However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.  For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.   -->

A major challenge in this work is inferring the sex of fossil specimens. This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020). Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).

We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G. doryssus is part of the extant threespine stickleback species complex (Gastersteus aculeatus).  Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.
We then combined these sets of data and treat the sex of the fossil specimens as missing and  used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.  Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.

After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes

# Data {#sec:data}
```{r}
stickle <- read_csv('stickle_revamped.csv')

@@

## Fossil Specimen Data
We used Gasterosteus doryssus data that were previously reported by @Stuart2020, @Voje2022, and Siddiqui et al. (in review). Briefly, the data were collected from fossil Series K from Quarry D (Cerasoni et al., in review), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA. Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years. Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1). Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete (Bell et al. 2006; Bell 2009; Stuart et al. 2020). This lineage subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (Bell et al. 2006; Stuart et al. 2020; Siddiqui et al. in review). The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (Hunt et al. 2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.

The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart. Table 2 shows the sample size at each of the 18 time periods.  There are at least 22 specimens at each time period with a high of 67 specimens in period 7.

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	sl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
\begin{table}[ht]
\centering
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
\label{tabn}

\end{table}

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendry2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (Baumgartner et al. 1988; @Schluter1992) (Table 2 (What is this?)). The generalist populations were collected by YES in 2013 and previously described in @Stuart2017. These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and his lab at University of British Columbia. They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018. The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.

We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirre2014. X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (Schindelin et al. 2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). All photographs, X-rays, and ObjectJ files have been uploaded to Morphosource.org (accession # TBD).
We dissected individuals from the generalist populations to determine sex from the gonads. Individuals from the species-pair lakes were sexed by Schluter and his lab, using a genotyping protocol (confirm and cite).

The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex.  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

## Outlier analysis. 
To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (pooled across lakes). We noted trait values greater than 3.5 standard deviations from the mean as potential outliers. We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were. We turned the remaining outlier trait values to NAs. (confirm)  (Wait, what?  Why?  Anything that was 3.5 SD above the mean was recorded as missing?  How do you justify this?  )

Missing data imputation, including fossil sex

Quantification and evolution of sexual dimorphism

@@

We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2009 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.  Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets. In addition to this, @ZhouReiter2009 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.  We proceed with our Bayesian analysis in this manner.

\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}

<!-- Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->

## Completed Data Model 
\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.   }

### Continuous Phenotypes
For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).

Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume
\begin{align}
{X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\
{X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align}
where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.  We further set \begin{align}
\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.  We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUProcess) where we assume $u_{k,gT}$ have a marginal mean of 0. Because each time period is ~1000 years, we will discretize the OU process without loss of generality.  For $t = 2,\ldots,T$, we assume
\begin{align}
u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
\label{eq:u_ar1}
\end{align}

@@

### Discrete Phenotypes
For the discrete phenotypes, we will assume
\begin{align}
{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
\label{eq:Y}
\end{align}
for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set
\begin{align}
\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
\label{eq:lambda}
\end{align}
We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume
\begin{align}
v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
\label{eq:v_ar1}
\end{align}
and
\begin{align}
v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
\label{eq:v1}
\end{align}

@@

# Discsusson, Future work and conclusions {#sec:conclusions}
We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback. For example, in lakes where sculpin competitors are absent and stickleback (Roesti et al. 2023)
See Spoljaric and Reimchen 2008, page 512 right column for references and discussion of differences between benthic males and limnetic females. Male stickleback are benthic and littoral (Wootton 1976)…. Reimchen papers in general good for this section.

@@

We thank O. Abughoush, S. Blaine, A. Chaudhary, M. Islam, F. Joaquin, C. Lawson-Weinert, R. Sullivan, J. Tien, M.P. Travis, and W. Shim for help with data collection. We thank D. Schluter and S. Blain for loaning specimens and sharing data. We thank K. Swagel and C. McMahan of the Field Museum for assistance with specimen x-rays. This research was supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the Center for Field Research (Earthwatch), and the National Geographic Society (2869-84) to MAB. It was also supported by NSF grants DEB-1456462 and EAR-2145830 to YES.  And NSF DMS-2015374 (GJM)
